### 2019-10-23 LCAforActiveSonarImaging/submission/final/lca_final.tex
2019-10-22

So far we have found it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, each steered within the range $|\phi|\in[0,\phi_\mathrm{3\,dB}(\beta)]$. What remains is to determine the number of windows needed. We study this in \Fig{oversampling_mosaic}, where we compared LCA images made from windows databases of different sizes, but with the same parameter boundaries. We included the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to an unsteered rectangularly weighted DAS.

Observe that sampling the $\beta$-range more finely increases the sidelobe suppression but leaves resolution unchanged. Adding more steering-variations improves both sidelobe suppression and resolution. We can observe a major improvement going to $N_\beta=2$ and $N_\phi=3$, but minimal improvement by adding yet more windows.

@@

We compare LCA image quality to that of DAS and MVDR in \Fig{image_quality}. In \Fig{beamformer_comparison_cross} we display the images computed by LCA using Kaiser windows, LCA using trigonometric windows, and MVDR. Two lateral image cuts through all images are presented in the left plot. The images and their corresponding cuts are nearly identical for the LCA version with Kaiser windows and the one with trigonometric windows. Compared to DAS the LCA produces targets with an FWHM that lies in between that of DAS and MVDR.

@@

The computational complexity of MVDR is generally considered to be of O($M^3$). However, the implementation we are using is optimized so that building the spatial covariance matrix is of O($N_kN_L,L^2$), and inverting it is of O($L^3$). We describe this implementation in \cite{Buskenes2014}, and a beamspace version in \cite{Asen2013}, where we optimize if for a graphics processing unit (GPU) for nearly two orders of magnitude speed increase compared to a straightforward C implementation. However, MVDR is not ideal for GPUs due to the complex data dependencies present in the covariance computation, and at best we only managed to utilize 10\% of the GPUs potential.

In comparison, LCA is of O($M N_\beta N_\phi$), and extremely well suited for GPUs. Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access. If we use a trigonometric window function we solve for the window parameter analytically as described in appendix \ref{sec:reducing_lca_complexity}. This reduces the complexity to O($M N_\phi$), and the only remaining LCA parameter is the steering amount.

@@

We found the LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser windows or trigonometric ones. However, LCA must be able to choose between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity. We achieved good results with 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$. For the steering angle we suggest a value in the range $\phi\in[\pm\delta\phi_\mathrm{3\,dB}(\beta)]$, where $\phi_\mathrm{3\,dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameter. Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively. Hence, this parameter can be used to control the aggressiveness and resolution gain of LCA.

In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able produce images similar to that of MVDR. Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using e.g. GPUs. Since LCA performs just as well if we use trigonometric windows instead of Kaiser ones, we can also compute the optimal window parameter analytically for any given steering angle, and avoid the search over the window parameter entirely.

@@

To control the resolution gain of LCA, we wanted to determine the extent that the windows need to be steered to cut the angle of the \minus{}3\,dB point of the window's spatial amplitude response by a fixed amount, say, by a factor 2 as reference. We computed this steering angle for some common configurations of the mentioned parameters. Refer to \Fig{steering_angle_3dB} for the result. The boundary values of the Kaiser window is shown, i.e. the rectangular window at $\beta=0$ and the near gaussian window at $\beta=5$, as well as one in between at $\beta=2.5$. Observe that for a given $M$ and $\beta$, the angle $\phi_\mathrm{3\,dB}$ scales near proportionally with $\frac{d}{\lambda}$. We found this to be true also for $\frac{d}{\lambda}\in\{0.5,4\}$. Hence, an approximate figure for a rather wide range of system parameters can be derived off this figure.

@@

This article is accompanied by some multimedia files and source code for determining the steering angles for LCA's windows. They are located at:

@@

\input{bio/jo_inge_buskenes}
\input{bio/roy_edgar_hansen}
\input{bio/andreas_austeng}

>>>>

So far we have found it reasonable to use Kaiser windows in the range $\beta\in[0,5]$, each steered within the range $|\phi|\in[0,\phi_\mathrm{3\,dB}(\beta)]$. What remains is to determine the number of windows needed. We study this in \Fig{oversampling_mosaic}, where we compare LCA images made from windows databases of different sizes, but with the same parameter boundaries. Included are the results from using a single window ($N_\beta=1$) and steering angle ($N_\phi=1$), which corresponds to an unsteered rectangularly weighted DAS.

Observe that a finer sampling of the $\beta$-range improves sidelobe suppression but leaves resolution unchanged. Adding more steering-variations improves both sidelobe suppression and resolution. We observe major improvement going to $N_\beta=2$ and $N_\phi=3$, but minimal improvement by adding yet more windows.

@@

We compare LCA image quality to that of DAS and MVDR in \Fig{image_quality}. In \Fig{beamformer_comparison_cross} we display the images computed by LCA using Kaiser windows, LCA using trigonometric windows, and MVDR. Two lateral image cuts through all images are presented in the left plot. The images and their corresponding cuts are nearly identical for the LCA version with Kaiser windows and the one with trigonometric windows. Compared to DAS the LCA produces images with an FWHM that lies in between that of DAS and MVDR.

@@

The computational complexity of MVDR is generally considered to be of O($M^3$). However, the implementation we are using is optimized so that building the spatial covariance matrix is of O($N_kN_L,L^2$), and inverting it is of O($L^3$). We describe this implementation in \cite{Buskenes2014}, and a beamspace version in \cite{Asen2013}, where we implement it on a graphics processing unit (GPU) for nearly two orders of magnitude speed increase compared to a straightforward C implementation. However, MVDR is not ideal for GPUs due to the complex data dependencies present in the covariance computation, and at best we only managed to utilize 10\% of the GPUs maximum theoretical potential.

In comparison, LCA is of O($M N_\beta N_\phi$), and extremely well suited for GPUs. Each pixel depends on only a small subset of data, and the windows can be precomputed and stored in GPU cache for near immediate access. Furthermore, if we use the trigonometric window function we can solve for the window parameter analytically as described in Appendix \ref{sec:reducing_lca_complexity}. This reduces the complexity to O($M N_\phi$), because the only remaining LCA parameter to create windows for will be the steering angle.

@@

We found LCA to be surprisingly insensitive to the exact window function it is used with, supported by the observation that near identical images were produced whether we used Kaiser windows or trigonometric ones. However, LCA must be able to choose between wide and narrow window responses, since the wide ones offer the best sidelobe suppression and the narrow ones offer the best sensitivity. We achieved good results with 2-3 Kaiser windows in the range $\beta\in[0,5]$, and observed minimal difference when adding more windows or adjusting the upper bound of $\beta$. For the steering angle we suggest a value in the range $\phi\in[\pm\delta\phi_\mathrm{3\,dB}(\beta)]$, where $\phi_\mathrm{3\,dB}(\beta)$ is the bandwidth of the respective window, and $\delta$ is a scaling parameter. Setting $\delta$ to either 50\% or 100\% reduces the lateral point target size to 73\% or 53\% compared to the rectangular window, respectively. Hence, this parameter can be used to control the aggressiveness and resolution gain of LCA.

In summary, LCA is an attractive alternative to other adaptive beamformers due to being very fast, simple to understand, practically parameter-free, inherently robust, and able to produce images similar to that of MVDR. Contrary to MVDR, it has a low computational complexity and can be easily and efficiently accelerated using e.g. GPUs. Also, since LCA performs just as well if we use trigonometric windows instead of Kaiser ones, we can compute the optimal window parameter analytically for any given steering angle and avoid the search over the window parameter entirely.

@@

To control the resolution gain of LCA, we wanted to determine the extent that the windows need to be steered to cut the angle of the \minus{}3\,dB point of the window's spatial amplitude response by a fixed amount, say, by a factor 2 as reference. We computed this steering angle for some common configurations of the mentioned parameters, see the result in \Fig{steering_angle_3dB}. The boundary values of the Kaiser window is shown, i.e. the rectangular window at $\beta=0$ and the near gaussian window at $\beta=5$, as well as one in between at $\beta=2.5$. Observe that for a given $M$ and $\beta$, the angle $\phi_\mathrm{3\,dB}$ scales near proportionally with $\frac{d}{\lambda}$. We found this to be true also for $\frac{d}{\lambda}\in\{0.5,4\}$. Hence, an approximate figure for a rather wide range of system parameters can be derived off this figure.

@@

This article is accompanied by some multimedia files and source code for determining steering boundaries for LCA's windows. The files are located at:

@@

\input{bio/jo_inge_buskenes_new}
\input{bio/roy_edgar_hansen_new}
\input{bio/andreas_austeng}

### 2019-10-23 LCAforActiveSonarImaging/submission/submit2/answer/answer.tex
2019-10-22

\begin{center}
\begin{tabular}{l p{.6\linewidth}}\centering
Manuscript number: & 2015JOE001805 \\
Title: & Low Complexity Adaptive Sonar Imaging \\
Authors: & Buskenes,~J.I., Hansen,~R.E., Austeng,~A.
\end{tabular}
\end{center}

>>>>

\begin{center}
\begin{tabular}{l p{.6\linewidth}}\centering
Manuscript number: & 2015JOE001805 \\
Title: & Low-Complexity Adaptive Sonar Imaging \\
Authors: & Buskenes,~J.I., Hansen,~R.E., Austeng,~A.
\end{tabular}
\end{center}

### 2019-10-23 ThesisJo/2_background.tex
2019-10-22

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers. After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig. \ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at an angle or point at a time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_mvdr.svg}}
\caption{MVDR (minimum variance distortionless response) beampattern is dynamic and continuous (adaptive beamforming). Interference suppression is good for prevalent noise sources assuming sufficient spatial statistics.}%
\label{2_fig_scenario_mvdr}
\end{figure}

@@

Element-space MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile. It seeks the weight set that minimizes noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
where $\dvec{a}$ is steering vector. Applying Lagrange multipliers to this convex optimization problem yields the solution
%
\begin{gather}
\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}} \qquad\text{where}\qquad \dvec{a}=\dvec{1}, \label{2_eq_weights}
\end{gather}
%
where $\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array. (\ref{2_eq_weights}) is the maximum likelihood estimate of the signal in the presence of zero-mean random Gaussian noise. We estimate $\R$ with a sample covariance matrix $\eR$, requiring some degree of
%
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Widrow1982,Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%

requiring some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors~\cite{Synnevag2009a}.

Combined, these techniques also ensure a numerically well conditioned $\eR$.

\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%

We split the array into subarrays of length $L$ that overlap on all but one sensor. If we let $x_l[n]$ represent the datavector from subarray $l$,
%
\begin{gather}
\dvec{x}_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}^T,
\end{gather}
%
then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as
%
\begin{gather}
\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{N_L-1}\sumb{n'=n-K}{n+K} \x_l[n]\x_l^H[n] \in\mathbb{C}^{L,L},\label{2_eq_spatialR}
\end{gather}
%
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays. In the actual design a fraction of the total power of $\eR[n]$ is finally added to its diagonal~\cite{Synnevag2007}, but this step will be ignored here due to its negligible contribution to runtime.

To summarize as in Figure \ref{mvdr_beamforming}: For every pixel in the image we estimate $\eR$ (\ref{2_eq_spatialR}), solve the linear equation $\eRi\1$ (\ref{2_eq_weights}), the finally substitute some to get the beamformer output (\ref{2_eq_mvdr_def}-\ref{2_eq_weights}).

>>>>

Consider an acoustic signal transmitted to highlight a surface of interest. Assume the backscattered wavefield is properly sampled using a uniform linear array (ULA) of receivers. After stripping the carrier frequency by heterodyning and the signal signature by matched filtering (Fig. \ref{2_fig_article_relation}), an image is formed by coherently combining the receiver outputs to focus at one position (or angle) at a time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_mvdr.svg}}
\caption{MVDR (minimum variance distortionless response) beampattern is dynamic and continuous (adaptive beamforming). Interference suppression is good for prevalent noise sources, assuming sufficient spatial statistics.}%
\label{2_fig_scenario_mvdr}
\end{figure}

@@

Element-space MVDR often improves image contrast and resolution compared to conventional beamformers~\cite{Benitz1997,Synnevag2007,Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}, but is computationally complex and statistically fragile. It seeks the weight set that minimizes noise and interference accumulated by the array, while enforcing unit gain for signals from the pixel of interest,
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
where $\dvec{a}$ is steering vector. Applying Lagrange multipliers to this convex optimization problem yields the solution
%
\begin{gather}
\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}}, \label{2_eq_weights}
\end{gather}
%
where $\dvec{a}=\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array. These weights yield the maximum likelihood estimate of the signal in the presence of zero-mean random Gaussian noise---but not in active systems typically characterized by highly correlated noise, interference and signal~\cite{Widrow1982}. For regular array geometries such as ours, this can be handled with spatial smoothing.

We estimate $\R$ with a sample covariance matrix $\eR$, requiring some degree of
%
\begin{itemize}
\item \emph{spatial smoothing} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal smoothing} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%
%
These techniques ensure a numerically well conditioned $\eR$~\cite{VanTrees2002,Kailath1985}, allow reducing the computational complexity and expose sufficient data parallelism for GPU-acceleration.

Formulation of the sample covariance matrix $\eR$ begins by splitting the array into subarrays of length $L$ that overlap on all but one sensor. If we let $x_l[n]$ represent the datavector from subarray $l$,
%
\begin{gather}
\dvec{x}_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}^T,
\end{gather}
%
then $\eR$ formed with temporal and spatial averaging (smoothing) is given as
%
\begin{gather}
\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{N_L-1}\sumb{n'=n-K}{n+K} \x_l[n]\x_l^H[n] \in\mathbb{C}^{L,L},\label{2_eq_spatialR}
\end{gather}
%
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays. Finally, a fraction of the total power of $\eR[n]$ is added to its diagonal~\cite{Synnevag2007}.

To summarize: For every pixel in the image we estimate $\eR$ (\ref{2_eq_spatialR}), solve the linear equation $\eRi\1$ (\ref{2_eq_weights}), the finally substitute some to get the beamformer output (\ref{2_eq_mvdr_def}-\ref{2_eq_weights}). Complexity is cubic with the number of sensors, $O(M^3)$, as opposed to linear for conventional beamformers, $O(M)$.

sample spatial spectral matrices
asymptotic behaviour
Forward-backward averaging
Toeplitz-structured covariance - assuming stationarity
Structured spectral matrix estimation
parametric spectral matrix estimation
singular value decomposition
fixed diagonal loading
